---
title: "Practical 01 — Intro to Linux"
---

## Goals

- Learn basic shell commands
- Practice editing files
- Practice loading data in R and Plotting

`ls` , `cd`, `mkdir`, `wget` (for plinks tool), `tar` or `unzip`,


### 1. List files
```bash
ls 
```

What else I can do with this command?
Options:
```bash
ls --help
```

```bash
man ls 
```
Note: press on **Q** keyboard button to **exit**

classify list (files or folders):
```bash
ls -F
```

list all:
```bash
ls -a
```

print the author of each file:
```bash
ls -la
```



### 2. Where am I?

What is my present working directory?

```bash
pwd
```


### 3. Create a folder/directory in current working directory

```bash
mkdir my_tools
```

multiple sub folders at once:
```bash
mkdir my_tools/plink
```
list them:
```bash
ls -a
```


### 4. Go in the new directory
```bash
cd my_tools
```
Or go to another sub directory
```bash
cd /plink
```


How do you know you moved in or not?

look at the texts before the `$` sign in the terminal
```bash
pwd
```
What if you type `cd ..` or `cd -` or `cd --`

How to come back to the place where you were?


### 5. Download the official Plink softwere binary and make it available
We use `wget` command for this 
see all version [here](https://www.cog-genomics.org/plink/)
```bash
wget https://s3.amazonaws.com/plink1-assets/plink_linux_x86_64_20250819.zip
```
#### Unzip the downloaded file
`unzip` command
```bash
unzip plink_linux_x86_64.zip
```

#### Make the binary executable: change the permision
```bash
chmod +x plink
```

#### Add Plink to your PATH 
(*optional but recommended*)
```bash
echo 'export PATH=$PATH:~/my_tools/plink' >> ~/.bashrc
source ~/.bashrc
```

#### Test installation
```bash
plink --version
```
You will see something printed as like as bellow: 

>PLINK v1.9.0-b.7.11 64-bit (19 Aug 2025)


## To do:

- Create a folder and a sub folder and 
- make a file in 1st folder there and 
- move it to the subfolder
- print if it is there
- make it executable

### 6. create a file and write something in it


```bash
nano test.txt
```

copy and paste following line

`This is a file 
we like to read it
and save it`

`Ctrl` + `right click` mouse

>`Ctrl` + `o` for write out, then `Enter` to save the name, then `Ctrl` + `x` for exit

## To do
- edit the new file and add the line 'Today is Monday' and exit the file or editing


### Word count
```bash
wc -la
```

#### print and pipe

```bash
cat | wc -l
```


## To do

- count no. of files or number of lines in a file text file
- see the couple of lines
- open the file with less
- use | pipe
- save output to a file


::: {.callout-tip title="Solution & explanation" collapse="true"}



#### 📁 File Setup
Download or place these files (`📁 data_4_linux`) in your working directory: 

- *example.vcf*: VCF file with genotype data
- *regions.bed*: BED file with genomic regions
- *plink_data.ped* and *plink_data.map*: Input files for Plink

#### 🔹 vcftools

1. Calculate allele frequencies
```bash
vcftools --vcf example.vcf --freq

```
- `--vcf example.vcf`: Specifies the input VCF file.
- `--freq`: Calculates allele frequencies for each variant.
- Output: `out.frq` (or use `--out` to name it).

2. Filter variants by minor allele frequency
```bash
vcftools --vcf example.vcf --maf 0.05 --recode --out filtered_maf
```
- `--maf` `0.05`: Filters out variants with MAF < 0.05.
- `--recode`: Outputs a new VCF file with only the filtered variants.
- `--out` `filtered_maf`: Prefix for output files.

3. Count missing data per individual
```bash
vcftools --vcf example.vcf --missing-indv --out missing_data

```
- `--missing-indv`: Calculates missing genotype rates per individual.
- Output: `missing_data.imiss`

#### 🔹 bcftools

1. View VCF header
```bash
bcftools view -h example.vcf

```
- `-h`: Displays only the header lines of the VCF file.

2. Filter variants by quality
```bash
bcftools view -i 'QUAL>30' example.vcf -o highqual.vcf

```
- `-i 'QUAL>30'`: Includes only variants with quality > 30.
- `-o` `highqual.vcf`: Output file name.

3. Convert VCF to BCF
```bash
bcftools view -O b -o output.bcf example.vcf
```
- `-O` `b`: Output format is BCF (binary VCF).
- `-o` `output.bcf`: Output file name.

#### 🔹 bedtools

1. Intersect VCF with BED regions
```bash
bedtools intersect -a example.vcf -b regions.bed > vcf_in_regions.vcf
```

- `-a`: First file (VCF) to compare.
- `-b`: Second file (BED) with regions of interest.
- `>`: Redirects output to a new file.

2. Count intersected variants
```bash
grep -v "^#" vcf_in_regions.vcf | wc -l
```

- `grep` `-v` `"^#"`: Removes header lines.
- `wc` `-l`: Counts the number of remaining lines (i.e., variants).

#### 🔹 Plink

1. Calculate allele frequencies
```bash
plink --file plink_data --freq --out plink_freq
```

- `--file` `plink_data`: Uses plink_data.ped and plink_data.map.
- `--freq`: Calculates allele frequencies.
- `--out` `plink_freq`: Output prefix.

2. Check missingness
```bash
plink --file plink_data --missing --out plink_missing
```
- `--missing`: Calculates missing genotype rates per individual and per SNP.
- Output: `.imiss` and `.lmiss` files.


:::



### Recommended packages 
Install **ggplot2** package

```r
install.packages("ggplot2")

library("ggplot2")
```

### Example R code

See the data structure:
```{r}
str(cars)

```
A data frame with 50 observations on 2 variables.

[,1]	speed	numeric	Speed (mph)
[,2]	dist	numeric	Stopping distance (ft)


### Plotting a figure
Speed and Stopping Distances of Cars
```{r}
# ggplot(data = cars, aes (x = speed, y = dist))+
# geom_point()
```

### Load data table in R 
### How to do that from RStudio?
### How to do that with function or command?
```{r}
# loading data set
myY=read.table(file="https://github.com/jiabowang/GAPIT/raw/refs/heads/master/Documents/mdp_traits.txt", head = TRUE)
```

